Reduced Nucleotide Diversity on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-08-08 manuscript.tex
edits, table1

\documentclass[9pt,twocolumn,twoside]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers

%\usepackage[colorinlistoftodos]{todonotes} % comments in margins
%\setlength{\marginparwidth}{1.25cm}
%\definecolor{flame}{rgb}{0.89, 0.35, 0.13}
%\newcommand{\josh}[1]{\textcolor{flame}{\emph{\scriptsize {#1}}}}
%\newcommand{\citex}{\textcolor{red}{\bf (CITE)\,}}
%\newcommand{\out}[1]{\textcolor{red}{#1}} %outline placeholder text
\newcommand{\X}{\textcolor{red}{\bf X\,}}

@@

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, both positive and negative selection can remove genetically linked neutral variation, resulting in a reduction in $N_{e}$ and the level of neutral polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, our analysis revealed that nucleotide diversity on the Y is approximately 12.5 fold lower than predicted from models of neutral evolution. We demonstrate that the magnitude of this reduction is unlikely to be explained by the observed occurrence of female biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations, we show that the diversity reduction on the Y is consistent with the predicted effects of background selection arising from suppressed recombination. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our results indicate that the low level of sequence diversity commonly observed on ancient Y chromosomes might evolve soon after sex chromosomes originate, with background selection playing an important role during the early stages of their evolution.

%this is 254 words and cannot be any longer than 250. Make edits accordingly.

%Add citations using the \verb|\citep{}| command, for example \citep{neher2013genealogies} or for multiple citations, \citep{neher2013genealogies, rodelsperger2014characterization}

\end{abstract}

@@

\maketitle
\thispagestyle{firststyle}
\marginmark
\firstpagefootnote
\correspondingauthoraffiliation{jhough@ucdavis.edu}
\vspace{-11pt}%

@@

In addition to demographic factors, evolutionary theory predicts that genetic variability on sex chromosomes should also be affected by the evolution of suppressed X-Y recombination, which lowers the coalescent $N_{e}$ for Y-linked genes and increases their vulnerability to the effects of selective sweeps \citep{smith1974hitch,aquadro1994} and background selection \citep{charlesworth1996background,charlesworth1994effect}. These processes can be viewed as examples of the Hill–Robertson (HR) effect, in which the effective population size for a given chromosomal region depends strongly on the rate of recombination, such that selection at a focal site is not independent of selection at nearby sites \citep{hill1966HReffect}. Moreover, both background selection and selective sweeps predict a reduction in $N_{e}$ and, as a result, reduced levels of neutral polymorphism, selection efficacy, and rate of adaptation \citep{comeron2008}. Although the HR effect can occur throughout the genome \citep{mcvean2000}, large genomic regions in which recombination is suppressed, such as the Y chromosome, are expected to experience the most severe reductions in neutral diversity due to chromosome-wide linkage \citep{charlesworth1996CB,charlesworth2000degeneration,bachtrog2013NRG}. Consistent with this, studies examining Y chromosome variability in mammals \citep{hellborg2004,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have found that Y-linked diversity is considerably lower than predicted under models of neutral evolution, suggesting that the HR effect plays a key role determining patterns of nucleotide diversity during sex chromosome evolution across a broad range of animal taxa.

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity and the forces contributing to sex chromosome divergence \citep{ellegren2011,otto2011PAR,bachtrog2013NRG}, we still know little about the influence of the HR effect on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. The timescales over which selection affects neutral polymorphism on sex chromosomes is therefore not well understood.

Recent studies of young sex chromosome system in plants have revealed that several canonical features of X and Y chromosome evolution are shared between plants and animals, including the genetic deterioration of the Y chromosome \citep{bergero2015,hough2014}, the occurrence of ‘evolutionary strata’ \citep{bergero2009}, and the evolution of dosage compensation \citep{muyle2012,papadopulos2015}. Nonetheless, except for a small number of genes in \textit{Silene latifolia} \citep{filatov2001diversity,qiu2010nucleotide}, the associated changes in genetic diversity that are predicted from evolutionary models of X- and Y-chromosome evolution have not been widely studied in young plant sex chromosomes. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}, studying recently evolved plant sex chromosomes provides an important test of the temporal dynamics of the HR effect.

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution, we analyzed single nucleotide polymorphisms (SNPs) on sex chromosomes and autosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with highly heteromorphic X and Y chromosomes that are estimated to have evolved  approximately 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making sex chromosomes in this species over 100 million years younger than the well-studied mammalian sex chromosomes \citep{lahn1999,ross2005dna}. \textit{R. hastatulus} has also received particular attention because of the unique occurrence of an intraspecific polymorphism for sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct chromosomal races \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species (the North Carolina race) is thought to have originated through an X-autosome fusion, with the XY system (the Texas race) maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, there is evidence that both the ancestral and neo-Y chromosomes have undergone significant genetic degeneration\citep{hough2014}.

Of particular relevance to the present study, \textit{R. hastatulus} populations throughout their geographical range (southern USA) have also been found to consistently exhibit female-biased sex ratios, with a mean sex ratio of $N_{m}/N_{f}=0.6$ \citep{pickup2013influence}. This is important because female-biased sex ratios are predicted to affect genetic diversity on sex chromosomes and autosomes \citep{ellegren2009}. Here, we investigated the factors affecting nucleotide diversity on X, Y, and autosomal chromosomes. Our goal was to determine whether levels of diversity on these chromosomes were jointly consistent with predictions from models of sex ratio bias and variance in reproductive success, or whether linked selection has been the dominant factor affecting patterns of neutral polymorphism on sex chromosomes.

\section*{Materials and Methods}
\subsection*{Population Samples and Sex-Linked Genes}
We analyzed sex-linked and autosomal genes identified from Illumina RNA sequence data from 24 individuals (12 males and 12 females, with 1 male and 1 female from each of 12 populations, 6 populations of each sex chromosome race). Samples of both sex chromosomal race were collected in 2010 from throughout the native range of \textit{R. hastatulus} (locations in Table S1), and plants were grown in the glasshouse from seeds collected from open-pollinated females. We extracted RNA from leaf tissue using Spectrum Plant Total RNA kits (Sigma-Aldrich). The isolation of mRNA and cDNA synthesis was conducted according to standard Illumina RNAseq procedures, with sequencing conducted on two Illumina HiSeq lanes with 150-bp end reads at the Genome Quebec Innovation Center. Reads from these 24 samples were mapped to the \textit{R. hastatulus} reference transcriptome \citep{hough2014}, and raw sequences are available from the Gen-Bank Short Read Archive under accession no. SRP041588. Reads were mapped using the Burrows–Wheeler Aligner \citep{li2010fast}, followed by Stampy\citep{lunter2011stampy}. We used Picard tools (http://picard.sourceforge.net) to modify mapping output into the format required for the Genome Analysis Toolkit \citep{mckenna2010genome} variant calling software, and subsequently removed genes with low coverage (<10x) and Phred Quality Scores <20. The population samples analyzed here were previously reported in Hough et al. (2014), where they were used to validate the ascertainment of sex-linked contigs identified through segregation analysis. From the list of sex-linked genes identified in Hough et al. (2014) that were shared between the Texas (XY) and North Carolina (XY1Y2) sex chromosome races, we identified a total of 460 sex-linked genes for analysis.

\subsection*{Autosomal Genes}
In evaluating evidence for nucleotide diversity differences between X and Y chromosomes, it is important to distinguish between reduced Y-linked diversity, and the possibility that X-linked diversity is elevated above the level predicted from a neutral model. To do this, we normalized sex-linked diversity estimates by autosomal diversity, and compared empirical X/A and Y/A nucleotide diversity ratios to those predicted from a neutral model (described below). Because the criteria for identifying autosomal loci in Hough et al. (2014) were based on the occurrence of four segregating SNPs per locus, this set of genes is probably higher in diversity than the average autosome. Here, we incorporated the broader set of all non-sex linked (punitively autosomal) genes from our transcriptome data. We filtered this set to remove those that may have been sex-linked but were not identified as such by the ascertainment criteria in Hough et al. (2014). In particular, we removed: (i) any genes in which there was evidence for at least one SNP with a sex-linked segregation pattern, (ii) any genes with SNPs showing fixed heterozygosity in males and fixed homozygosity in females, (iii) any genes with less than 10X coverage or greater than 100X coverage from independently obtained genomic coverage data (to filter out duplicate genes or those with highly repetitive sequences), and (iv) any genes containing SNPs with large (>0.4) allele frequency differences between males and females. Finally, we removed genes with fewer than 100 synonymous sites following this filtering to avoid biasing our results toward genes that may have been particularly short due to assembly problems. This filtering resulted in a set of 12,356 and 11,350 autosomal genes in the Texas and North Carolina races, respectively.

@@

>>>>

\documentclass[9pt,twocolumn,twoside]{gsajnl}
% Use the documentclass option 'lineno' to view line numbers
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{geometry} % just not to bother with the table width
%\usepackage[colorinlistoftodos]{todonotes} % comments in margins
%\setlength{\marginparwidth}{1.25cm}
%\definecolor{flame}{rgb}{0.89, 0.35, 0.13}
%\newcommand{\josh}[1]{\textcolor{flame}{\emph{\scriptsize {#1}}}}
%\newcommand{\citex}{\textcolor{red}{\bf (CITE)\,}}
%\newcommand{\out}[1]{\textcolor{red}{#1}} %outline placeholder text
\newcommand{\X}{\textcolor{red}{\bf X\,}}

@@

\begin{abstract}
X and Y chromosomes differ in effective population size ($N_{e}$), rates of recombination, and exposure to natural selection, all of which can significantly affect levels of genetic variability within the genome. On Y chromosomes with suppressed recombination, both positive and negative selection can remove genetically linked neutral variation, resulting in a reduction in $N_{e}$ and the level of polymorphism on Y compared to X chromosomes or autosomes. However, non-selective factors including biased sex ratios and high variance in male fitness can also reduce Y-linked $N_{e}$, making it difficult infer the causes of low Y-chromosome diversity. Here, we investigate the factors affecting X- and Y-linked polymorphism during the early stages of plant sex chromosome evolution in \textit{Rumex hastatulus} (Polygonaceae). Strikingly, we find that nucleotide diversity on the Y chromosome is 40 fold lower than on the X chromosome. We demonstrate that the magnitude of this reduction is not consistent with a purely neutral model based on the observed occurrence of female biased sex ratios in this species, or by reduced male $N_{e}$ caused by high variance in male fitness. Rather, using forward simulations, we show that the diversity reduction on the Y is consistent with the predicted effects of purifying selection removing deleterious mutations and linked neutral variation. Given the recent origin of \textit{R. hastatulus} sex chromosomes, our results indicate that the low level of sequence diversity commonly observed on ancient Y chromosomes might evolve soon after sex chromosomes originate, with background selection playing an important role during the early stages of their evolution.
\end{abstract}

@@

\maketitle
\thispagestyle{firststyle}
\marginmark
\firstpagefootnote
\correspondingauthoraffiliation{jhough@ucdavis.edu}
\vspace{-11pt}

@@

In addition to demographic factors, evolutionary theory predicts that genetic variability on sex chromosomes should be affected by the evolution of suppressed X-Y recombination, which lowers the coalescent $N_{e}$ for Y-linked genes and increases their vulnerability to the effects of selective sweeps \citep{smith1974hitch,aquadro1994} and background selection \citep{charlesworth1996background,charlesworth1994effect}. These processes can be viewed as examples of the Hill–Robertson (HR) effect, in which the effective population size for a given chromosomal region depends strongly on the rate of recombination, such that selection at a focal site is not independent of selection at nearby sites \citep{hill1966HReffect}. Moreover, both background selection and selective sweeps predict a reduction in $N_{e}$ and, as a result, reduced levels of neutral polymorphism, selection efficacy, and rate of adaptation \citep{comeron2008}. Although the HR effect can occur throughout the genome \citep{mcvean2000}, large genomic regions in which recombination is suppressed, such as the Y chromosome, are expected to experience the most severe reductions in neutral diversity due to chromosome-wide linkage \citep{charlesworth1996CB,charlesworth2000degeneration,bachtrog2013NRG}. Consistent with this, studies examining Y chromosome variability in mammals \citep{hellborg2004,Wilsonsayres2014} and Drosophila \citep{mcallister1999,bachtrog2000} have found that Y-linked diversity is considerably lower than predicted under models of neutral evolution, suggesting that the HR effect plays a key role determining patterns of nucleotide diversity during sex chromosome evolution across a broad range of animal taxa.

Despite widespread interest in determining the effects of recombination on patterns of genetic diversity \citep{ellegren2011,bachtrog2013NRG}, we still know very little about the influence of the HR effect on patterns nucleotide polymorphism in younger sex chromosome systems, where \textit{de novo} recombination suppression evolved relatively recently \citep{charlesworth2016plant}. Recent studies of young plant sex chromosome systems have revealed that several canonical features of X and Y chromosome evolution are shared between plants and animals, including the genetic deterioration of the Y chromosome \citep{bergero2015,hough2014}, the occurrence of ‘evolutionary strata’ \citep{bergero2009}, and the evolution of dosage compensation \citep{muyle2012,papadopulos2015}. Nonetheless, except for a small number of genes in \textit{Silene latifolia} \citep{filatov2001diversity,qiu2010nucleotide}, the associated changes in genetic diversity that are predicted from evolutionary models of X- and Y-chromosome evolution have not been widely studied in plant sex chromosomes. Given that background selection and selective sweeps are predicted to have the strongest effects during the earliest stages of Y-chromosome evolution, before Y-chromosomes have lost the majority of their genes \citep{bachtrog2008temporal}, studying recently evolved sex chromosomes provides an important test of the temporal dynamics of the HR effect.

To investigate the factors affecting nucleotide diversity in the early stages of sex chromosome evolution, we analyzed single nucleotide polymorphisms (SNPs) on sex chromosomes and autosomes in the plant \textit{R. hastatulus }(Polygonaceae). This species is a dioecious annual with highly heteromorphic X and Y chromosomes that are estimated to have evolved  approximately 15 MYA \citep{quesada2011,grabowska2015,navajas2005}, making sex chromosomes in this species over 100 million years younger than the well-studied mammalian sex chromosomes \citep{lahn1999,ross2005dna}. \textit{Rumex hastatulus} has also received particular attention because of the unique occurrence of an intraspecific sex chromosome system, in which both XY and $XY_{1}XY_{2}$ males occur in geographically distinct 'chromosomal races' \citep{smith1963mechanism}. The $XY_{1}XY_{2}$ sex chromosome system in this species (the North Carolina race) is thought to have originated through an X-autosome fusion, with the XY system (the Texas race) maintaining the ancestral chromosome complement \citep{smith1964evolving}. Despite the recent origin of sex chromosomes in both races, a recent study revealed that both the ancestral and neo-Y chromosomes have undergone significant genetic degeneration\citep{hough2014}.

Of particular relevance to the present study, \textit{R. hastatulus} populations have also been found to consistently exhibit female-biased sex ratios, with a mean sex ratio of $N_{m}/N_{f}=0.6$ \citep{pickup2013influence}. This is important because female-biased sex ratios are predicted to affect chromosomal $N_{e}$ and therefor the level of genetic diversity maintained on sex chromosomes and autosomes \citep{ellegren2009}. Here, we investigated whether the patterns of diversity we observed on sex chromosomes and autosomes were jointly consistent with predictions from models of sex ratio bias and variance in reproductive success, or whether linked selection (background selection and selective sweeps) has played a greater role.

\section*{Materials and Methods}
\subsection*{Population Samples and Sex-Linked Genes}
We analyzed sex-linked and autosomal genes identified from Illumina RNA sequence data from 24 individuals (12 males and 12 females, with 1 male and 1 female from each of 12 populations; 6 populations of each sex chromosome race). Samples were collected in 2010 from throughout the native range of \textit{R. hastatulus} (locations in Table S1), and plants were grown in the glasshouse from seeds collected from open-pollinated females. We extracted RNA from leaf tissue using Spectrum Plant Total RNA kits (Sigma-Aldrich). The isolation of mRNA and cDNA synthesis was conducted according to standard Illumina RNAseq procedures, with sequencing conducted on two Illumina HiSeq lanes with 150-bp end reads at the Genome Quebec Innovation Center. Reads from these 24 samples were mapped to the \textit{R. hastatulus} reference transcriptome \citep{hough2014}, and raw sequences are available from the Gen-Bank Short Read Archive under accession no. SRP041588. Reads were mapped using the Burrows–Wheeler Aligner \citep{li2010fast}, followed by Stampy \citep{lunter2011stampy}. We used Picard tools (http://picard.sourceforge.net) to modify mapping output into the format required for the Genome Analysis Toolkit \citep{mckenna2010genome} variant calling software, and subsequently removed genes with low coverage (<10x) and Phred Quality Scores <20. The population samples analyzed here were previously reported in Hough et al. (2014), where they were used to validate the ascertainment of sex-linked genes identified through segregation analysis. Here, we focused on the set of 460 sex-linked genes for which the Y homolog was inferred to be on the $y_{1}$ chromosome (i.e., sex-linked genes that are shared between the Texas and North Carolina races).

\subsection*{Autosomal Genes}
In evaluating evidence for nucleotide diversity differences between X and Y chromosomes, it is important to distinguish between reduced Y-linked diversity, and the possibility that X-linked diversity is elevated above the level predicted from a neutral model. To do this, we normalized sex-linked diversity estimates by autosomal diversity, and compared empirical X/A and Y/A nucleotide diversity ratios to those predicted from a neutral model (described below). Because the criteria for identifying autosomal loci in Hough et al. (2014) were based on the occurrence of four segregating SNPs per locus, this set of genes is probably higher in diversity than the average autosome. Therefore, for the present analysis we incorporated the broader set of all non-sex linked (punitively autosomal) genes from our transcriptome data. We also filtered this set to remove genes that may have been sex-linked but were not identified as such by the conservative ascertainment criteria in Hough et al. (2014). In particular, we removed: (i) any genes in which there was evidence for at least one SNP with a sex-linked segregation pattern, (ii) any genes with SNPs showing fixed heterozygosity in males and fixed homozygosity in females, (iii) any genes with less than 10X coverage or greater than 100X coverage from independently obtained genomic coverage data (to filter out duplicate genes or those with highly repetitive sequences), and (iv) any genes containing SNPs with large (>0.4) allele frequency differences between males and females. Finally, we removed genes with fewer than 100 synonymous sites following this filtering to avoid biasing our results toward genes that may have been particularly short due to assembly problems. This filtering resulted in a set of 12,356 and 11,350 autosomal genes in the Texas and North Carolina races, respectively.

@@

\begin{table}
\centering
\caption{Estimates of neutral diversity by race on \textit{R. hastatulus} sex chromosomes and autosomes.}
\begin{tabular}{ccccccccc}
\textbf{} & \multicolumn{2}{c}{\textbf{Texas}} & \multicolumn{2}{c}{\textbf{South Carolina}} & \multicolumn{2}{c}{\textbf{Florida}} \\
chromosome & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ & $\theta$ & $\theta/\theta_{A}$ \\
\midrule
A & 0.006 & 1 & 0.006 & 1 & 0.005 & 1 \\    
X & 0.0047 & 0.85 & 0.002 & 0.33 & 0.0047 & 0.37 \\ 
Y & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 & $\num{e-4}$ & 0.002 \\ 
\addlinespace

\bottomrule
\end{tabular}

\end{table}
